---
title: "Group Project 1: Peanut Butter vs Almond Butter Analysis"
subtitle: "DSST389: Advanced Data Science"
author: "*Anush Margaryan & Cooper Woods*"
institute: "University of Richmond"
date: 2/17/2025
format:
  html:
    embed-resources: true
  pdf: default
editor: source
toc: true
---

## Introduction

Approximately **4.5 million U.S. adults** have peanut allergies, making it one of the most prevalent food allergies. You can read more about peanut allergy prevalence and characteristics [here](https://www.jacionline.org/article/S0091-6749%2820%2932412-X/fulltext). Thus, it is important to find **nutritionally comparable alternatives**.

For our project, we investigate whether almond butter, a popular alternative, can provide **similar protein content** to peanut butter. Additionally, we compare the **degree of processing (FPro)** characterizing peanut and almond butter preparation techniques. **It is important to note that a lower FPro score indicates that the food is less processed.**

We obtained our data from [True Food](https://www.truefood.tech), powered by a comprehensive database Grocery DB, for peanut and almond butter **protein content** and **FPro score** across the available brands.

### Library Setup
Below we will import the necessary libraries for our analysis.
```{r}
#| label: libraries
#| echo: true
#| warning: false

library(tidyverse) #Data manipulation, visualization, and analysis
library(ggrepel) #Improved readability of labels
library(dplyr) #Efficient data manipulation
library(stringr) #Character string manipulation
library(rstatix) #Used for calculating error bars
theme_set(theme_minimal()) #Minimalistic theme for the plots
```

## About the data
We begin by loading the dataset. Here are the first few rows of `butters`:

```{r}
#| echo: false

butters <- read_csv("../data/butters.csv") #Read the dataset
head(butters) #Display the first few rows
```
### Description

- The `butters` data was extracted from the **True Food** database.
- It contains **32 rows**, each representing 11 peanut and 21 almond butter brands.
  
### Data dictionary

Columns we will analyze include:

| Column Name | Description                     | Unit                 |
|-------------|---------------------------------|----------------------|
| `name`      | Product name as marketed        | Character (text)     |
| `fpro_score`| Food processing score           | Scaled 0 to 1        |
| `protein`   | Protein content per serving     | Grams (g)            |
| `category`  | Type of butter                  | "Peanut" or "Almond" |

## Data Analysis and Visualizations

```{r}
#| echo: false
# Renaming the 'category'column to 'Type of Butter' for analysis clarity

butters <- butters |>
  rename(`Type of Butter` = category) 
```

### Scatterplot of FPro Score vs. Protein Conent (g)

```{r}
#| label: protein-vs-fpro-scatterplot
#| fig-cap: "Figure 1. Scatterplot showing the relationship between food processing score (FPro) and protein content per serving (g) for various peanut (in orange) and almond (in brown) butters."
#| echo: false

butters |>
  ggplot(aes(x = fpro_score, y = protein,
             color = `Type of Butter`)) +
  geom_point() +
  geom_smooth(aes(color = `Type of Butter`),
      method = "lm",
      se = FALSE,
      formula = y ~ x) +
  geom_text_repel(aes(label = name), size = 2.5) +
  labs(title = "Almond Butter vs Peanut Butter Analysis",
       x = "Food Processing Score (FPro)",
       y = "Protein Amount (g)"
       ) +
    scale_color_manual(values = c("almond butter" = "chocolate4", "peanut butter" = "darkorange"),
                       labels = c("almond butter" = "Almond", "peanut butter" = "Peanut"))
```

### Bar Chart Comparing Average FPro Scores for Almond and Peanut Butters

```{r}
#| label: fpro-score-bar-chart
#| fig-cap: "Figure 2. Bar Chart comparing the protein content per serving (g) for various peanut (in orange) and almond (in brown) butters. 
#| The standard error bars show variability within each butter type."
#| echo: false

butters_summary <- butters |>
  group_by(`Type of Butter`) |>
  summarise(
    mean_fpro = mean(fpro_score, na.rm = TRUE),
    se_fpro = sd(fpro_score, na.rm = TRUE) / sqrt(n()) #calculating the standard error
  ) 
ggplot(butters_summary, aes(x = `Type of Butter`, y = mean_fpro, fill = `Type of Butter`)) +
  geom_bar(stat = "identity", position = position_dodge(), color = "black") +
  geom_errorbar(aes(ymin = mean_fpro - se_fpro, 
                    ymax = mean_fpro + se_fpro),
                position = position_dodge(0.9),
                width = 0.3 #determining the minimum and maximum ranges of the standard error
                ) +
  scale_fill_manual(values = c("almond butter" = "chocolate4", "peanut butter" = "darkorange"),
                       labels = c("almond butter" = "Almond", "peanut butter" = "Peanut")) +
  labs(title = "Almond Butter vs Peanut Butter Analysis",
       x = "Butter Type",
       y = "Food Processing Score (FPro)")
```
## Conclusions

Overall, our analysis suggests that there is no clear correlation between processing and protein content. While **peanut butter** has naturally **higher protein content**, **almond butter** seems to be generally **less processed**. Thus, individuals with peanut allergies should look for a balance bwteen lower processing and higher protein content in almond butters.


