---
title: "Lenon vs. McCartney: Textual Analysis of Beatles Lyrics"
subtitle: "DSST389: Advanced Data Science"
author: "*Cooper Woods*"
institute: "University of Richmond"
date: 4/23/2025
echo: true
format:
  html:
    anchor-sections: true
    code-tools: false
    code-link: true
    embed-resources: true
    pdf: default
    link-external-icon: true
    link-external-newwindow: true
    number-sections: true
    smooth-scroll: true
    toc: true
editor:
  markdown:
    wrap: 72
---

## Introduction

This project compares the lyrical styles of John Lennon and Paul McCartney by analyzing the text of Beatles songs attributed to each songwriter. Using topic modeling and sentiment analysis, the goal is to uncover themes and emotional tones unique to the main two writers of the Beatles.

## Loading the Libraries and Dataset

```{r}
#| label: libraries
#| output: FALSE

# Load necessary libraries
library(ggrepel)
library(tidytext)
library(textdata)
library(scales)
library(tidyverse)
library(topicmodels)
library(glue)
library(tm)
set.seed(123)
theme_set(theme_minimal())

packages <- c(
  "tidyverse", "tidymodels", "textrecipes", "kknn",
  "xgboost", "glmnet", "corrplot", "parallel", "future",
  "textdata", "stm", "topicmodels", "tidytext", "reshape2",
  "ldatuning", "tm", "glue"
)

for (pkg in packages) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    install.packages(pkg, dependencies = TRUE)
  }
}
for (pkg in packages) {
  library(pkg, character.only = TRUE)
}
```

```{r}
#| label: q-set-up-file
#| echo: false
#| warning: false

beatles <- read_csv("../data/beatles_data.csv") |>
  rename(
    song = "Song",
    album_debut = "Album Debut",
    songwriter = "Songwriter(s)",
    lead = "Lead Vocal(s)",
    year = "Year",
    lyrics = "Lyrics"
  ) |>
  select(song, album_debut, songwriter, lead, year, lyrics)

head(beatles)
```

## Prepare and Tokenize the Data

```{r}
#| label: q-tokenize
#| output: FALSE

tidy_lyrics <- beatles |>
  unnest_tokens(input = lyrics, output = word)
```

## Filter for Lennon and McCartney Songs

```{r}
#| label: song-filter

tidy_lyrics_filtered <- tidy_lyrics |>
  filter(songwriter %in% c("Lennon", "McCartney")) |>
  distinct(song, word) |>
  count(word) |>
  filter(n > 1) |>
  inner_join(tidy_lyrics, by = "word") |>
  filter(songwriter %in% c("Lennon", "McCartney"))
```

## Topic Modeling by Songwriter

```{r}
#| label: topic-modeling
#| #| label: song-filter
#| message: false

# Prepare documents
song_docs <- beatles |>
  filter(songwriter %in% c("Lennon", "McCartney")) |>
  group_by(songwriter, song) |>
  summarize(text = glue_collapse(lyrics, sep = " "), .groups = "drop")

# Create DTM per songwriter
lennon_dtm <- song_docs |>
  filter(songwriter == "Lennon") |>
  unnest_tokens(word, text) |>
  anti_join(get_stopwords()) |>
  count(song, word) |>
  cast_dtm(document = song, term = word, value = n)

mccartney_dtm <- song_docs |>
  filter(songwriter == "McCartney") |>
  unnest_tokens(word, text) |>
  anti_join(get_stopwords()) |>
  count(song, word) |>
  cast_dtm(document = song, term = word, value = n)

# Fit LDA separately using 5 terms. I'm choosing 5 because that number isn't too large or too small
lennon_lda <- LDA(lennon_dtm, k = 5, control = list(seed = 123))
mccartney_lda <- LDA(mccartney_dtm, k = 5, control = list(seed = 123))

# Extract and plot top terms for Lennon
lennon_terms <- tidy(lennon_lda, matrix = "beta") |>
  group_by(topic) |>
  slice_max(beta, n = 5) |>
  ungroup()

lennon_terms |>
  mutate(term = fct_reorder(term, beta)) |>
  ggplot(aes(x = term, y = beta, fill = factor(topic))) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~topic, scales = "free") +
  coord_flip() +
  labs(
    title = "Top Terms for Each Lennon Topic",
    x = NULL, y = "Beta"
  )

# Extract and plot top terms for McCartney
mccartney_terms <- tidy(mccartney_lda, matrix = "beta") |>
  group_by(topic) |>
  slice_max(beta, n = 5) |>
  ungroup()

mccartney_terms |>
  mutate(term = fct_reorder(term, beta)) |>
  ggplot(aes(x = term, y = beta, fill = factor(topic))) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~topic, scales = "free") +
  coord_flip() +
  labs(
    title = "Top Terms for Each McCartney Topic",
    x = NULL, y = "Beta"
  )
```

These plots present the top terms in five topics identified for each
songwriter, revealing distinctive thematic elements in Lennon and
McCartney's lyrical styles.

I interpret Lennon's topics as follows:

1. Vulnerability and emotional songs  
2. Possible relational storytelling  
3. Focus on female subject / love interest  
4. Love and longing  
5. Philosophical / political messaging

I interpret McCartney's topics as follows:

1. Personal / story-based songs  
2. Most likely tied to the album, *Sgt. Pepper's Lonely Hearts Club Band*, more surreal songs  
3. Romantic / love songs  
4. Hard to tell, maybe about moving on  
5. Repetitive songs about characters, including *Hey Jude* and *Paperback Writer*

As a whole, it would appear that Lennon leans introspective,
philosophical, and emotionally raw, where McCartney embraces
storytelling and character-driven lyrics with more emotional clarity.
The topic model reinforces well-known stylistic differences between the
two writers and highlights the diversity within each songwriter’s
catalog.

## Sentiment Analysis

```{r}
#| label: sentiment-analysis

sentiment_scores <- tidy_lyrics_filtered |>
  inner_join(get_sentiments("afinn"), by = "word") |>
  group_by(song, songwriter) |>
  summarize(sentiment = mean(value, na.rm = TRUE), .groups = "drop")
```

```{r}
#| label: sentiment-plot

sentiment_scores |>
  ggplot(aes(x = songwriter, y = sentiment, fill = songwriter)) +
  geom_violin(trim = FALSE, alpha = 0.7) +
  geom_boxplot(width = 0.1, fill = "white") +
  labs(
    title = "Sentiment of Lennon vs. McCartney Songs",
    y = "Average Sentiment Score", x = NULL
  ) +
  scale_fill_manual(values = c("Lennon" = "#FFD700", "McCartney" = "#1E90FF")) +
  theme(legend.position = "none")
```

The violin plot shows the distribution of sentiment scores for songs by
each songwriter. McCartney's songs show slightly higher and more
consistent positive sentiment, while Lennon's have more variation and
dip into negative territory more often. Boxplots embedded in the violins
confirm that the median sentiment is higher for McCartney, and Lennon’s
interquartile range is broader.

## Sentiment Trends Over Time

```{r}
#| label: sentiment-over-time

sentiment_by_year <- sentiment_scores |>
  inner_join(select(beatles, song, year), by = "song") |>
  filter(year <= 1970) |>
  group_by(year, songwriter) |>
  summarize(avg_sentiment = mean(sentiment), .groups = "drop")

sentiment_by_year |>
  ggplot(aes(x = year, y = avg_sentiment, color = songwriter)) +
  geom_line(linewidth = 1) +
  geom_point() +
  scale_color_manual(values = c("Lennon" = "#FFD700", "McCartney" = "#1E90FF")) +
  labs(
    title = "Sentiment Trends Over Time",
    y = "Average Sentiment", x = "Year"
  )
```

This plot shows how the average sentiment of Lennon and McCartney’s
Beatles-era (1962-1970) songs fluctuated over time.

-   **Early Years (1962–1964)**: Both songwriters start off relatively
    positive. McCartney’s sentiment spikes in 1964, possibly reflecting
    upbeat hits during the height of Beatlemania.

-   **Mid 60s (1965–1967)**: Sentiment begins to dip, especially sharply
    for Lennon in 1966 and McCartney in 1967. These years align with
    more experimental and introspective songwriting (*Revolver*, *Sgt.
    Pepper's*), and real-world pressures like touring fatigue and
    personal evolution.

-   **Late 60s (1968–1969)**: Sentiment converges and levels off.
    Lennon’s work becomes slightly more positive again in 1967, before
    **both songwriters land in a more emotionally neutral range** by
    1969 — possibly mirroring rising tensions within the band and more
    mature lyrical themes.

## Conclusion

This project reveals how **John Lennon** and **Paul McCartney** used language in their own way. Lennon's lyrics show greater *emotional variability and* intensity, while McCartney's are more *consistently upbeat and structured*. The *sentiment trends over time* suggest that both writers evolved with the band's changing tone, yet largely stayed true to their own styles. *Topic modeling* reinforces this - Lennon's songs often circle themes of *existential tension and personal struggle*, while McCartney's songs focus on *narrative, nostalgia, and emotional clarity*.

These differences show how **individual voice shapes collective art**. Quantifying lyrical tendencies helps us see familiar music in a new way, representing artistic identity in measurable language patterns.
